---
title: Beads community trait data
author: "Masatoshi Katabuchi"
date: "`r format(Sys.time(), '%B %d, %Y')`"
fontsize: 12pt
format:
  html:
    theme: cosmo #readable #sandstone #spacelab #flatly
    # highlight: pygments #tango #kate
    toc: true
    toc-depth: 2
    number-sections: true
    smooth-scroll: true
    embed-resources: false
---

```{r, include=FALSE}
library(tidyverse)
library(picante)
library(factoextra)
library(mvtnorm)
```


# Phylo

```{r}
tr <- read.tree("data-raw/qjzad.tre")
sp <- tr$tip.label
tr_dist <- vcv(tr) |> cov2cor()

phy_tr <- rmvnorm(2, rep(0, ncol(tr_dist)), tr_dist)

tmp <- tibble(tmp = phy_tr[1, ], sp = colnames(tr_dist)) |>
  arrange(tmp)
```

## Mass artifact and phylogenetic signal

```{r}
set.seed(123)
d <- read_csv("data/nature02403-s2.csv", skip = 10) |>
  janitor::clean_names()

data <- d |>
  dplyr::select(sp = species, log_ll, log_lma, log_amass, log_nmass, log_pmass, log_rdmass) |>
  filter(!is.na(log_lma)) |>
  mutate(across(starts_with("log_"), ~ . * log(10)))

log_lma <- data |>
  pull(log_lma)

lma <- exp(log_lma)

data |>
  pull(log_amass) |>
  exp() |>
  summary()
```


Dummy LL is based on the regression of LL on LMA.

```{r}
fit <- lm(log_ll ~ log_lma, data = data)
sd_log_ll <- sd(data$log_ll, na.rm = TRUE)

# coef(fit)[1]
ll <- rlnorm(length(lma),
             coef(fit)[1] + coef(fit)[2] * log_lma - sd_log_ll^2 / 2,
             sd_log_ll)
```


Dummy A_mass is based on Model-A in Osnas et al. 2013.

$$
X_{MiK} = \bar{X_{Ai}} LMA_k^{-1} \epsilon_{Aik}
$$

For Amass,
$\bar{X_{Ai}}$ = 11642 and $\epsilon_{Aik}$ = 0.55


```{r}
a_mass <- rlnorm(length(lma),
               log(11642  / lma) - 0.55^2 /2,
               0.55)
```

I don't add area-based traits in the data, but they can be calculated as follows:

$$
X_{AiK} = X_{Mik} LMA_k.
$$


Other LES traits are also based on Model-A.

```{r}
r_mass <- rlnorm(length(lma),
               log(1402  / lma) - 0.5^2 /2,
               0.5)

n_mass <- rlnorm(length(lma),
               log(1.94  / lma) - 0.46^2 /2,
               0.46)

p_mass <- rlnorm(length(lma),
               log(0.1305  / lma) - 0.6^2 /2,
               0.6)
```

Wood density and seed mass are independent with the LES.

```{r}
wd <- rlnorm(length(lma),
             log(0.6) - 0.2^2 / 2., 0.2)

sm <- rlnorm(length(lma),
             log(2.3) - 1^2 / 2., 1)
```

Add the phylogenetic signal to the above data.

```{r}
data2 <- tibble(
  lma,
  ll,
  a_mass,
  r_mass,
  n_mass = n_mass * 100,
  p_mass = p_mass * 100,
  wd,
  sm) |>
  round(2) |>
  sample_n(length(sp)) |>
  arrange(lma)

dummy_trait <- bind_cols(tmp, data2) |>
  arrange(-sm)

# Find the rows containing the target species
row_henryi <- which(dummy_trait$sp == "Lithocarpus_henryi")
row_xylocarpus <- which(dummy_trait$sp == "Lithocarpus_xylocarpus")
row_wattii <- which(dummy_trait$sp == "Castanopsis_wattii")

# Swap the top row with Lithocarpus_henryi
dummy_trait[c(1, row_henryi), 2] <- dummy_trait[c(row_henryi, 1), 2]

# Swap the second row with Lithocarpus_xylocarpus
dummy_trait[c(2, row_xylocarpus), 2] <- dummy_trait[c(row_xylocarpus, 2), 2]

# Swap the third row with Castanopsis_wattii
dummy_trait[c(3, row_wattii), 2] <- dummy_trait[c(row_wattii, 3), 2]

dummy_trait <- dummy_trait |>
  arrange(sp)

summary(dummy_trait)

```

Units are:

- LMA (g m^-2^)
- LL (months)
- *A*~mass~ (nmol g^-1^ s^-1^)
- *R*~mass~ (nmol g^-1^ s^-1^)
- *N*~mass~ (%)
- *P*~mass~ (%)
- Wood density (g cm^-3^)
- Seed mass (mg)

```{r}
dummy_trait |>
  DT::datatable()
```

```{r}
# Select the top and bottom species for 'sm' and 'lma' simultaneously
lma_sp <- dummy_trait %>%
  arrange(lma) %>%
  slice(c(1:5, (n()-4):n())) %>%
  pull(sp)
sm_sp <- dummy_trait %>%
  arrange(sm) %>%
  slice(c(1:5, (n()-4):n())) %>%
  pull(sp)

# Extract unique and sorted species names in one step
abundant_sp <- unique(c(lma_sp, sm_sp)) %>%
  sort()

abundant_sp
```

# Check PCA

```{r}
pca_res <- prcomp(log(data2), scale = TRUE)

fviz_pca_biplot(pca_res,
                axis = c(1, 2),
                axes.linetype = "dotted",
                col.var = "black",
                alpha.var = 0.9,
                pointsize = 2.5,
                alpha.ind = 0.1,
                geom.ind = "point",
                palette = c("#7570b3", "#1b9e77"),
                addEllipses = TRUE,
                ellipse.level = 0.95,
                title = "",
                arrowsize = 0.5,
                #labelsize = 2,
                label = "all",
                mean.point = FALSE,
                repel = TRUE)
```

```{r}
tr2 <- chronos(tr, lambda=0)
plot(tr)
plot(tr2)
```


# Check k signal

```{r}
tr2 <- chronos(tr, lambda=0)
tr3 <- multi2di(tr2, random=TRUE)
#tr$edge.length <- tr$edge.length + rnorm(length(tr$edge.length), 0, 0.01)

physig_tb <- tibble()
for (i in 3:10) {
  x <- dummy_trait[,i] %>% unlist
  names(x) <- dummy_trait$sp
  physig_tb  <- bind_rows(physig_tb, phylosignal(x[tr3$tip.label], tr3))
}

physig_tb2 <- physig_tb %>%
  mutate(trait = names(dummy_trait)[3:10])

physig_tb2

```


# Save

```{r}
dummy_trait %>%
  dplyr::select(-tmp) |>
  arrange(sp) |>
  write_csv("data/dummy_trait.csv")

tr3 |>
  write.tree("data/dummy_tree.newick")
```

# Environment

```{r}
devtools::session_info()
```


```{r}
knitr::knit_exit()
```



```{r}
summary(a_mass)

data |>
  pull(log_amass) |>
  exp() |>
  summary()

tmp <- data |>
  pull(log_amass)

summary(10^tmp)

data |>
  mutate(log_lma_inv = -log_lma) %>%
  lm(log(a_mass) ~ log_lma_inv, data = .)
```

```{r}
dummy_trait2 <- dummy_trait |>
  arrange(lma)

dummy_trait3 <- bind_cols(tmp, dummy_trait2)

dummy_trait3 |>
  DT::datatable()
```
